The Courtroom Challenge\\\large Can Any Open-World AI Thrive Without Paying the Validation Bill?\\[1em]\small\textit{This note poses a conjecture, not a leaderboard: automate validation, prove it's unnecessary, or price human judgment honestly.}

### 2025-06-06 main.tex
Add files via upload
v2: edits after internal review

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=0.9\linewidth]{courtroom}%
      % no .pdf extension needed if LaTeX finds courtroom.pdf
  \caption{One step of the Courtroom loop.%
           \ Solid arrows: primary query flow.%
           \ Dashed arrows: helper calls and cost logging.%
           \ The cost\,bus aggregates all resource writes into a single ledger.}
  \label{fig:courtroom}
\end{figure}

@@

\section{Minimal State Machine}
\label{sec:minimal-state-machine}
Formal state $S_t=(q_t,h_t,a_t,M_t)$, validator $V_\theta$, memory update
$U$, cost ledger $\WriteCost = \sum_i \kappa_i\,c_i$.

%------------------------------------------------
\section{Formal Framework}
\label{sec:framework}
%------------------------------------------------

@@

  \item \textbf{Validator family.}  
  \(V_{\theta} : (Q, H, M) \to \{\textsc{ACCEPT}, \textsc{REJECT}, \textsc{ESCALATE}\}\), with a
  tunable validation–preference dial \(\theta \in \Theta\).
  \(V_{\theta}\) may be
  (i) a stronger ensemble LLM,
  (ii) a formal-proof kernel,
  (iii) a deterministic numeric or symbolic tool-chain that certifies or refutes the claim, or
  (iv) a human panel as fallback.\footnote{\textit{Self-play precedent.}  
  AlphaZero's value network functions as an automated judge for Monte-Carlo rollouts.
  The Courtroom validator is its open-domain analogue: any module that can reliably
  downgrade faulty claims can occupy the judge slot.}

@@

Given $S_t=(q_t,h_t,a_t,M_t)$:
\begin{enumerate}
  \item $h_t \leftarrow G(q_t,M_t)$
  \item $d_t \leftarrow V_\theta(q_t,h_t,M_t)$
        \begin{itemize}
          \item If $d_t=\mathrm{ACCEPT}$ set $a_t=h_t$
          \item Else obtain corrected answer $a_t = C(q_t,h_t)$
                (oracle $C$ may call a human; cost is logged)
        \end{itemize}
  \item $M_{t+1} \leftarrow U(M_t,q_t,h_t,a_t)$
  \item External stream provides $q_{t+1}$
\end{enumerate}

@@

Let $c_{i,t}$ be the measured consumption of resource $i$ at step~$t$
(bytes written, FAISS rebuild time, human minutes, \emph{etc.}).
Fix weights $\kappa_i$ once per benchmark—should the conjecture survive first contact.  The per-write cost is
\[
  \WriteCost_t \;=\; \sum_i \kappa_i\,c_{i,t}.
\]
Cumulative friction budget up to step $T$ is
$F_T = \sum_{t=0}^{T-1}\WriteCost_t$.

@@

>>>>

% --- macros used in the information-theoretic bound ---
\usepackage{amsmath,amssymb}   % (already loaded in many templates)

\newcommand{\Err}{\mathrm{Err}}
\newcommand{\Prb}{\mathbb{P}}  % avoid \Pr: clashes with TeX primitive
\newcommand{\kappabar}{\bar{\kappa}}
\newcommand{\E}{\mathbb{E}}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=0.9\linewidth]{courtroom}%
      % no .pdf extension needed if LaTeX finds courtroom.pdf
  \caption{One step of the Courtroom loop.%
           \ Solid arrows: primary query flow.%
           \ Dashed arrows: helper calls and cost logging.%
           \ The cost\,bus aggregates all resource writes into a single ledger.
           \ On \textsc{DEFER} the round is skipped (no cost, no memory write).}
  \label{fig:courtroom}
\end{figure}

@@

\section{Minimal State Machine}
\label{sec:minimal-state-machine}
Formal state $S_t=(q_t,h_t,a_t,M_t)$, validator $V_\theta$, memory update
$U$, cost ledger $\WriteCost = \sum_i \kappa_i\,c_i$.
%
Validator actions:
$\{\textsc{ACCEPT},\textsc{REJECT},\textsc{ESCALATE},\textsc{DEFER}\}$.
%------------------------------------------------
\section{Formal Framework}
\label{sec:framework}
%------------------------------------------------

@@

  \item \textbf{Validator family.}  
  \(V_{\theta} : (Q, H, M) \to \{\textsc{ACCEPT}, \textsc{REJECT}, \textsc{ESCALATE}, \textsc{DEFER}\}\), with a
  tunable validation–preference dial \(\theta \in \Theta\).
  \(V_{\theta}\) may be
  (i) a stronger ensemble LLM,
  (ii) a formal-proof kernel,
  (iii) a deterministic numeric or symbolic tool-chain that certifies or refutes the claim, or
  (iv) a human panel as fallback.\footnote{\textit{Self-play precedent.}  
  AlphaZero's value network functions as an automated judge for Monte-Carlo rollouts.
  The Courtroom validator is its open-domain analogue: any module that can reliably
  downgrade faulty claims can occupy the judge slot.} \textsc{DEFER} means “skip this round”: no answer is returned,
  no escalation cost is paid, and the generator proceeds to the next
  query.

@@

Given $S_t=(q_t,h_t,a_t,M_t)$:
\begin{enumerate}
  \item $h_t \leftarrow G(q_t,M_t)$
  \item $d_t \leftarrow V_\theta(q_t,h_t,M_t)$
        \begin{itemize}
          \item If $d_t=\mathrm{ACCEPT}$ set $a_t=h_t$
          \item Else obtain corrected answer $a_t = C(q_t,h_t)$
                (oracle $C$ may call a human; cost is logged)
          \item If $d_t=\textsc{DEFER}$ set $a_t=\bot$
                and skip Steps 3–4 below.
        \end{itemize}
  \item $M_{t+1} \leftarrow U(M_t,q_t,h_t,a_t)$
  \item External stream provides $q_{t+1}$
\end{enumerate}

@@

Let $c_{i,t}$ be the measured consumption of resource $i$ at step~$t$
(bytes written, FAISS rebuild time, human minutes, \emph{etc.}).
Fix weights $\kappa_i$ once per benchmark—should the conjecture survive first contact.  The per-write cost is
\[
  \WriteCost_t \;=\; \sum_i \kappa_i\,c_{i,t}.
\]
Cumulative friction budget …  $F_T = \sum_{t=0}^{T-1}\WriteCost_t$.%
\footnote{\textsc{DEFER} incurs zero cost, so the corresponding
$\WriteCost_t$ entry is~0.}

@@

\subsection{Information–theoretic lower bound}
\label{sec:info-bound}

\begin{lemma}[Interactive-Fano corollary]\label{lem:ErrLower}
  Fix a validator $V$ run for $T$ rounds under margin parameter $\delta>0$.
  Let
  \[
    \kappabar
      :=\mathbb{E}_{\omega}\bigl[\kappa(\omega)\bigr],
    \qquad
    \Err_{\mathrm{bar}}
      :=\mathbb{E}_{\omega}\bigl[\Err(\omega)\bigr].
  \]
  Suppose the transcript $X$ satisfies
  $I(\theta;X)\le I$
  under a uniform prior on an $M$-point label set ($M\ge 2$).
  Then
  \begin{equation}\label{eq:ErrLower}   % <-- new label
    \boxed{%
      \Err_{\mathrm{bar}}
      \;\;\ge\;\;
      \frac{\delta\,\kappabar\,T}{4}\;
      \Bigl(
        1-\dfrac{I+\log 2}{\log M}
      \Bigr)
    }.
  \end{equation}
\end{lemma}

\begin{proof}[Sketch]

\textbf{Classical identification bound.}
Choosing the 0–1 loss $L(M,X)=\mathbf 1\{\hat M(X)\neq M\}$
and threshold $\Delta=\tfrac12$
in Proposition 3 of \citet{chen2024interfano} yields
\[
  P_{\mathrm{id}}
  \;\ge\;
  \tfrac12\Bigl(1-\tfrac{I+\log 2}{\log M}\Bigr).
\]

\medskip
\noindent
To obtain a bound on label error we apply the same proposition to
the indicator
\[
  L^\star \;=\; \mathbf 1_{\{\Err>\Delta_{\mathrm{Err}}\}},
  \qquad
  \Delta_{\mathrm{Err}} := \tfrac12\,\delta\,\kappabar\,T.
\]

\item
  Re-apply Proposition 3 to
  $L^\star = \mathbf 1\{\Err>\Delta_{\mathrm{Err}}\}$
  with threshold $\Delta=\tfrac12$.
  Because $\log \!\bigl(1/\rho_{1/2}\bigr)\ge\log M$ for the uniform
  prior, the proposition yields the lower bound
  \[
    \Prb\!\bigl[\Err>\Delta_{\mathrm{Err}}\bigr]
      \;\ge\;
      \frac12\Bigl(
        1-\frac{I+\log 2}{\log M}
      \Bigr).
  \]

\item
  Multiply this probability by
  $\Delta_{\mathrm{Err}}=\tfrac12\,\delta\,\kappabar\,T$
  to bound the expectation:
  \[
    \E[\Err]
      \;\ge\;
      \Delta_{\mathrm{Err}}\;
      \Prb[\Err>\Delta_{\mathrm{Err}}]
      \;=\;
      \frac{\delta\,\kappabar\,T}{4}\;
      \Bigl(
        1-\frac{I+\log 2}{\log M}
      \Bigr).
      \qedhere
  \]

\end{proof}

\paragraph{Discussion.}
Inequality~\eqref{eq:ErrLower} separates the lower bound into two
factors:

\begin{enumerate}[label=(\roman*)]
\item the \emph{size factor} $\displaystyle\frac{\delta\,\kappabar\,T}{4}$,
      which grows linearly with the horizon~$T$ and is large whenever
      the validator answers many queries ($\kappabar$ moderate) under a
      tight margin~$\delta$;
\item the \emph{information gap}
      $\displaystyle\Bigl(1-\tfrac{I+\log 2}{\log M}\Bigr)$,
      which is positive as soon as the transcript carries fewer bits
      than the label space can encode.
\end{enumerate}
Together they formalise our narrative: \emph{a low-cost, low-friction
validator cannot keep cumulative label error small indefinitely as the
stream length~$T$ grows}, unless it either (a) escalates a non-negligible
fraction of rounds (making $\kappabar$ small and thus incurring cost) or
(b) relaxes the margin~$\delta$.  In particular, when $M$ is
exponentially large and $I$ is bounded by design, $\Err_{\mathrm{bar}}$
grows linearly with~$T$.
